---
title: "Your Paper Title Here"
author: "Your Name(s)"
date: Date
output: pdf_document
urlcolor: blue
header-includes:
- \usepackage{amsmath,amsfonts,amssymb}
- \usepackage{setspace} \doublespacing
fontsize: 11pt
---

```{r setup, include=FALSE}
#Use this code chunk to include libraries, and set global options.
```

# Abstract
Write your abstract here.

***TESTTTING****

# Keywords
*Keyword 1*, *Keyword 2*, *Keyword 3*, *Keyword 4*

\newpage

# Introduction / Background
Provide an introduction with background information.

```{r label="important_R_code", eval=TRUE, echo=FALSE}
# You can reference your code in the appendix (sample here).
```

# Statistical Methods

## Model
Describe the statistical model used.

## Likelihood Inference
Detail the likelihood approach.

## Bayesian Inference
Detail the Bayesian approach.

# Results
Present your findings. 

# Discussion / Conclusion
Discuss / conclude here.

# Bibliography

Brown, B. (2024). *Lecture Title*. Lecture slides, Course Name, University Name.

Doe, J. (2020). Title of the Paper. *Journal Name*, 12(3), 45-67.

Last, F., & Last, F. (2025). *Book Title*. Publisher.

Smith, A., & Johnson, C. (2023). *Title of the Online Article*. Retrieved from https://www.example.com.

# Appendix

## Code
```{r ref.label = "important_R_code", eval=FALSE}
```

## Proofs
If applicable, include detailed mathematical derivations or additional theoretical explanations.

Proof of the Maximum Likelihood Estimator for $\widehat{\psi}_0^{mle}$

We can begin with the Likelihood function. Recall that the design of our experiment is binomial; we observe 170 infections and try to make inference on $\pi$, the proportion of infections from those who have the vaccine. Assuming each observation is independent (which is necessary for a binomial model), we can write our likelihood function in terms of $\pi$:
$$L(\pi) = \binom{170}{x}(\pi)^x(1-\pi)^{170-x}$$
Where $x$ is the observed number of infections where the patient had the vaccine. We observe $x = 8$, so we can rewrite our likelihood function:
$$L(\pi; x = 8) = \binom{170}{8}(\pi)^8(1-\pi)^{162}$$
Notably, however, we are not hoping to draw inference on $\pi$, but rather on $\psi$, the efficacy of the vaccine. We can write this parameter in terms of $\pi$.
$$\psi = \frac{1-2\pi}{1-\pi}$$
In order to write our Likelihood function in terms of $\psi$, we can rewrite $\psi$ in terms of $\pi$:
$$\pi = \frac{\psi - 1}{\psi -2}$$
Now we rewrite our likelihood function.
$$L^*(\psi) = \binom{170}{8}\left(\frac{\psi-1}{\psi-2}\right)^8\left(\frac{-1}{\psi-2}\right)^{162}$$
Now that we have written our full likelihood function in terms of $\psi$, we can 
