---
title: "Your Paper Title Here"
author: "Your Name(s)"
date: Date
output: pdf_document
urlcolor: blue
header-includes:
- \usepackage{amsmath,amsfonts,amssymb}
- \usepackage{setspace} \doublespacing
fontsize: 11pt
---

```{r setup, include=FALSE}
#Use this code chunk to include libraries, and set global options.
```

# Abstract
Write your abstract here.

***TESTTTING****

# Keywords
*Keyword 1*, *Keyword 2*, *Keyword 3*, *Keyword 4*

\newpage

# Introduction / Background
Provide an introduction with background information.

```{r label="important_R_code", eval=TRUE, echo=FALSE}
# You can reference your code in the appendix (sample here).
```

# Statistical Methods

## Model
Describe the statistical model used.

## Likelihood Inference
Detail the likelihood approach.

#likihood function:

#MLE

#Confidence Interval for Phi

  #Large sample confidence interval
  
  #Bootstrap
  
```{r, echo= false}
B= 100
set.seed(123)
n <- 160
boot_df <- replicate ( n = B, expr= (rbinom(n = 1,size = n, prob = 8/170)))
boot_df

psi <- numeric(B)
for (i in 1:B){
  Tvar <- boot_df[i]
  psi[i] <- ((n-(2*Tvar)) / (n-Tvar) )
}                       
        
mean(psi)
psi

result <- quantile(psi, probs = c(0.025, 0.975))

# Print the result
print(result)

```
  

## Bayesian Inference
Detail the Bayesian approach.

#beta binomial model
$T_{\pi} \sim Binom(170,\pi)$
$g(\pi) = Beta(a,b)$

$P(\psi\geq 0.3) = P(\frac{1-2\pi}{1-\pi} \geq 0.3)$
$=P(\pi \leq \frac{7}{17})$
$=\frac{7}{17}$

#priors
We will use the following priors

experimental prior
$P(\psi \geq 0.3) = 0.700102$
$P(\psi \geq 0) = 1$

0.05 prior
$P(\psi \geq 0.3) = 0.05$
$P(\psi \geq 0) = 0.5$

0.01 prior
$P(\psi \geq 0.3) = 0.01$
$P(\psi \geq 0) = 0.5$

The experimental prior is the prior that Pfizer used in their study, which gives the vaccine the best chance of passing the tests. The following two priors have a decreasing chance of passing the tests.

We can visualize how much more likely their prior is than ours
```{r}

```


#betas

```{r, echo = false}
library(LearnBayes)
prior1Beta <- beta.select(quantile1 = list(p=0.05, x = 7/17),
            quantile2 = list(p = 0.5, x = 0.5))
prior2Beta <- beta.select(quantile1 = list(p=0.01, x = 7/17),
            quantile2 = list(p = 0.5, x = 0.5))
```

The Pfizer study used $Beta(0.700102,1)$ as their prior Betas.
We will be trying $Beta($`r round(prior1Beta,4)`$)$ and $Beta($`r round(prior2Beta,4)`$)$ as our prior Beta values.
So, the posterior distributions are as follows:
Pfizer: $h(\pi|t) = Beta($ `r 0.700102 + 8`, `r 162+0.700102`$)$
0.05 posterior: $h(\pi|t) = Beta($ `r round(prior1Beta[1],4) + 8`, `r 162+round(prior1Beta[1],4)`$)$
0.01 posterior: $h(\pi|t) = Beta($ `r round(prior2Beta[1],4) + 8`, `r 162+round(prior2Beta[1],4)`$)$

summarizing the posterior distribution of $\pi$

```{r}
posterior_median_pi <- qbeta(0.5, shape1 = 43.04 + 104, shape2 = 293 - 104 + 43.04 )
cat("posterior median", round(posterior_median_pi, 4))
qbeta(p = c(0.025, 0.975), shape1 = 43.04 + 104, shape2 = 293 - 104 + 43.04)
library(HDInterval)
hdi(qbeta, credMass=0.95, shape1 = 43.04 + 104,
shape2 = 293 - 104 + 43.04 )

```

# Results
Present your findings. 

# Discussion / Conclusion
Discuss / conclude here.

# Bibliography

Brown, B. (2024). *Lecture Title*. Lecture slides, Course Name, University Name.

Doe, J. (2020). Title of the Paper. *Journal Name*, 12(3), 45-67.

Last, F., & Last, F. (2025). *Book Title*. Publisher.

Smith, A., & Johnson, C. (2023). *Title of the Online Article*. Retrieved from https://www.example.com.

# Appendix

## Code
```{r ref.label = "important_R_code", eval=FALSE}
```

## Proofs
If applicable, include detailed mathematical derivations or additional theoretical explanations.

Proof of the Maximum Likelihood Estimator for $\widehat{\psi}_0^{mle}$

We can begin with the Likelihood function. Recall that the design of our experiment is binomial; we observe 170 infections and try to make inference on $\pi$, the proportion of infections from those who have the vaccine. Assuming each observation is independent (which is necessary for a binomial model), we can write our likelihood function in terms of $\pi$:
$$L(\pi) = \binom{170}{x}(\pi)^x(1-\pi)^{170-x}$$
Where $x$ is the observed number of infections where the patient had the vaccine. We observe $x = 8$, so we can rewrite our likelihood function:
$$L(\pi; x = 8) = \binom{170}{8}(\pi)^8(1-\pi)^{162}$$
Notably, however, we are not hoping to draw inference on $\pi$, but rather on $\psi$, the efficacy of the vaccine. We can write this parameter in terms of $\pi$.
$$\psi = \frac{1-2\pi}{1-\pi}$$
In order to write our Likelihood function in terms of $\psi$, we can rewrite $\psi$ in terms of $\pi$:
$$\pi = \frac{\psi - 1}{\psi -2}$$
Now we rewrite our likelihood function.
$$L^*(\psi) = \binom{170}{8}\left(\frac{\psi-1}{\psi-2}\right)^8\left(\frac{-1}{\psi-2}\right)^{162}$$
Now that we have written our full likelihood function in terms of $\psi$, we can 
