---
title: "Your Paper Title Here"
author: "Your Name(s)"
date: Date
output: pdf_document
urlcolor: blue
header-includes:
- \usepackage{amsmath,amsfonts,amssymb}
- \usepackage{setspace} \doublespacing
fontsize: 11pt
---

```{r setup, include=FALSE, warning=FALSE, echo = FALSE}
#Use this code chunk to include libraries, and set global options.
library(ggplot2)
library(HDInterval)
library(dplyr)
library(binom)
knitr::opts_chunk$set(
  echo = FALSE,      # show code in the output
  warning = FALSE,  # hide warnings
  message = FALSE   # hide messages
)
```

```{r label = "psi_dist", echo = FALSE}
psi_dist <- function(x, shape1, shape2){
  pi <- (x-1)/(x-2)
  return(dbeta(x = pi, shape1 = shape1, shape2 = shape2))
}
```

# Abstract
Write your abstract here.

***TESTTTING****

# Keywords
*Keyword 1*, *Keyword 2*, *Keyword 3*, *Keyword 4*

\newpage

# Introduction / Background
Provide an introduction with background information.

```{r label="important_R_code", eval=TRUE, echo=FALSE}
# You can reference your code in the appendix (sample here).
```

```{r}
# Summary data from the paper:
# For participants without evidence of infection:
# Group       Cases    SampleSize
# BNT162b2      8       17,411
# Placebo     162       17,511

data <- data.frame(
  Group = c("BNT162b2", "Placebo"),
  Cases = c(8, 162),
  SampleSize = c(17411, 17511)
)

# Calculate incidence rate per 1000 participants:
data <- data %>%
  mutate(IncidenceRate = Cases / SampleSize * 1000)

# Calculate exact binomial 95% confidence intervals for the incidence rates
ci <- binom.confint(data$Cases, data$SampleSize, methods = "exact")
data$Lower <- ci$lower * 1000
data$Upper <- ci$upper * 1000

# Create the bar plot
ggplot(data, aes(x = Group, y = IncidenceRate, fill = Group)) +
  geom_bar(stat = "identity", width = 0.6) +
  geom_errorbar(aes(ymin = Lower, ymax = Upper), width = 0.2) +
  labs(title = "Incidence Rate of Covid-19 Cases (per 1000 participants)",
       y = "Incidence Rate (per 1000)",
       x = "Group") +
  theme_minimal() +
  theme(legend.position = "none")

```
```{r}
# Calculate attack rates:
attack_rate_vaccine <- 8 / 17411
attack_rate_placebo <- 162 / 17511

# Vaccine efficacy (VE) is defined as 1 - (attack rate in vaccine group) / (attack rate in placebo group)
VE <- 1 - attack_rate_vaccine / attack_rate_placebo
VE_percent <- round(VE * 100, 2)

cat("Estimated Vaccine Efficacy (VE):", VE_percent, "%\n")

```

# Statistical Methods

## Model
Describe the statistical model used.

## Likelihood Inference
Detail the likelihood approach.

#likihood function:

#MLE

#Confidence Interval for Psi

  #Large sample confidence interval
  
  #Bootstrap
  
  We will also use a computational approach to find our $\hat\psi_{MLE}$ interval by bootstrapping our $\hat\pi$. We will randomly draw with replacement many times from a $Binomial(n, \pi_{observed})$ distribution. We will then calculate a value of $\hat\psi_i$ for each $\hat\pi_i$ from our bootstrap. We will then a construct a 95% confidence interval by calculating the 2.5th and 97.5th percentile of our $\hat\psi$s
  
  

## Bayesian Inference
Detail the Bayesian approach.
We will use a beta binomial model:

$T_{\pi} \sim Binom(170,\pi)$
$g(\pi) = Beta(a,b)$

$P(\psi\geq 0.3) = P(\frac{1-2\pi}{1-\pi} \geq 0.3)$
$=P(\pi \leq \frac{7}{17})$
$=\frac{7}{17}$

for our model we will use the following priors: 

experimental prior
$P(\psi \geq 0.3) \approx 0.57$
$P(\psi \geq 0) = 1$

0.05 prior
$P(\psi \geq 0.3) = 0.05$
$P(\psi \geq 0) = 0.5$

0.01 prior
$P(\psi \geq 0.3) = 0.01$
$P(\psi \geq 0) = 0.5$

The experimental prior is the prior that Pfizer used in their study, which gives the vaccine the best chance of passing the tests. The following two priors have a decreasing chance of passing the tests.



# Results
Present your findings. 




Our bootstrap 95% confidence interval is:

```{r, echo= FALSE}
B= 1000
set.seed(123)
n <- 160
boot_df <- replicate ( n = B, expr= (rbinom(n = 1,size = n, prob = 8/170)))
psi <- numeric(B)
for (i in 1:B){
  Tvar <- boot_df[i]
  psi[i] <- ((n-(2*Tvar)) / (n-Tvar) )
}                       

result <- quantile(psi, probs = c(0.025, 0.975))

# Print the result
print(result)

```


#Baysian Approach




```{r, echo = FALSE}
library(LearnBayes)
prior1Beta <- beta.select(quantile1 = list(p=0.05, x = 7/17),
            quantile2 = list(p = 0.5, x = 0.5))
prior2Beta <- beta.select(quantile1 = list(p=0.01, x = 7/17),
            quantile2 = list(p = 0.5, x = 0.5))
```

The Pfizer study used $Beta(0.700102,1)$ as their prior Betas.
We will be trying $Beta($`r round(prior1Beta,4)`$)$ and $Beta($`r round(prior2Beta,4)`$)$ as our prior Beta values.

We can visualize how much more likely their prior is than ours
  
```{r, echo = FALSE}

ggplot() +
geom_function(fun = psi_dist,
mapping = aes(color = "Pfizer"),
args = list(shape1 = 0.700102, shape2 = 1),
xlim = c(0,1)) +
geom_function(fun = psi_dist,
mapping = aes(color = "prior1"),
args = list(shape1 = 43.04, shape2 = 43.04),
xlim=c(0,1)) +
geom_function(fun = psi_dist,
mapping = aes(color = "prior2"),
args = list(shape1 = 85.63, shape2 = 85.63),
xlim=c(0,1)) +
scale_color_manual(name = "dist", 
                   values = c("Pfizer" = "blue", "prior1" = "red", "prior2" = "black"))+
labs(title = "Prior of experiment and new pirors",
x=expression(pi),
y = "PDF")
```

So, the posterior distributions are as follows:
Pfizer: $h(\pi|t) = Beta($ `r 0.700102 + 8`, `r 162+1`$)$
0.05 posterior: $h(\pi|t) = Beta($ `r round(prior1Beta[1],4) + 8`, `r 162+round(prior1Beta[1],4)`$)$
0.01 posterior: $h(\pi|t) = Beta($ `r round(prior2Beta[1],4) + 8`, `r 162+round(prior2Beta[1],4)`$)$

We can visualize the comparison between all of the priors and posteriors:

```{r, echo = FALSE}

ggplot() +
geom_function(fun = psi_dist,
mapping = aes(color = "Pfizer"),
args = list(shape1 = 0.700102, shape2 = 1),
xlim = c(0,1)) +
geom_function(fun = psi_dist,
mapping = aes(color = "Pfizer"),
args = list(shape1 = 0.700102 + 8, shape2 = 162 + 1),
xlim=c(0,1)) +
geom_function(fun = psi_dist,
mapping = aes(color = "0.05"),
args = list(shape1 = 43.04, shape2 = 43.04),
xlim = c(0,1)) +
geom_function(fun = psi_dist,
mapping = aes(color = "0.05"),
args = list(shape1 = 43.04 + 8, shape2 = 162 + 43.04),
xlim=c(0,1)) +
geom_function(fun = psi_dist,
mapping = aes(color = "0.01"),
args = list(shape1 = 85.63, shape2 = 85.63),
xlim = c(0,1)) +
geom_function(fun = psi_dist,
mapping = aes(color = "0.01"),
args = list(shape1 = 85.63 + 8, shape2 = 162 + 85.63),
xlim=c(0,1)) +
scale_color_manual(name = "dist", 
                   values = c("Pfizer" = "blue", "0.05" = "red", "0.01" = "black"))+
labs(title = "Priors and Posteriors",
x=expression(pi),
y = "PDF")


```


summarizing the posterior distribution of $\pi$

```{r, echo = FALSE}
# Pfizer Posterior
pfizer_shape1 <- 0.700102 + 8
pfizer_shape2 <- 162 + 1
pfizer_median <- qbeta(0.5, shape1 = pfizer_shape1, shape2 = pfizer_shape2)
pfizer_ci <- qbeta(c(0.025, 0.975), shape1 = pfizer_shape1, shape2 = pfizer_shape2)

# 0.05 Prior Posterior
prior1_shape1 <- prior1Beta + 8
prior1_shape2 <- 162 + prior1Beta
prior1_median <- qbeta(0.5, shape1 = prior1_shape1, shape2 = prior1_shape2)
prior1_ci <- qbeta(c(0.025, 0.975), shape1 = prior1_shape1, shape2 = prior1_shape2)

# 0.01 Prior Posterior
prior2_shape1 <- prior2Beta + 8
prior2_shape2 <- 162 + prior2Beta
prior2_median <- qbeta(0.5, shape1 = prior2_shape1, shape2 = prior2_shape2)
prior2_ci <- qbeta(c(0.025, 0.975), shape1 = prior2_shape1, shape2 = prior2_shape2)

cat("Pfizer posterior median:", round(pfizer_median, 4), "\n")
cat("95% CI for Pfizer: [", round(pfizer_ci[1], 4), ",", round(pfizer_ci[2], 4), "]\n\n")

cat("0.05 prior posterior median:", round(prior1_median, 4), "\n")
cat("95% CI for 0.05 prior: [", round(prior1_ci[1], 4), ",", round(prior1_ci[2], 4), "]\n\n")

cat("0.01 prior posterior median:", round(prior2_median, 4), "\n")
cat("95% CI for 0.01 prior: [", round(prior2_ci[1], 4), ",", round(prior2_ci[2], 4), "]\n")


```
  
## Significance Tests and Posterior Distributions

The Pfizer study used a $\text{Beta}(0.700102,1)$ prior. In our analysis, we compare this with two alternative prior distributions:
$$
\text{Beta}\Bigl(\text{Prior 1: } \text{Beta}\Bigl(`r round(prior1Beta[1],4)`,\, `r round(prior1Beta[1],4)`\Bigr) \Bigr)
\quad \text{and} \quad
\text{Beta}\Bigl(\text{Prior 2: } \text{Beta}\Bigl(`r round(prior2Beta[1],4)`,\, `r round(prior2Beta[1],4)`\Bigr) \Bigr).
$$

Given the data (with 8 successes and 162 failures), the posterior distributions are as follows:
  

- **Pfizer Posterior:**
$$
h(\pi\mid t) = \text{Beta}\Bigl(0.700102 + 8,\, 162 + 1\Bigr)
$$
  

- **0.05 Prior Posterior:**
$$
h(\pi\mid t) = \text{Beta}\Bigl(`r round(prior1Beta[1],4)` + 8,\, 162 + `r round(prior1Beta[1],4)`\Bigr)
$$

- **0.01 Prior Posterior:**
$$
h(\pi\mid t) = \text{Beta}\Bigl(`r round(prior2Beta[1],4)` + 8,\, 162 + `r round(prior2Beta[1],4)`\Bigr)
$$
  

Below we compute the posterior medians and approximate 95\% credible intervals (using the 2.5th and 97.5th percentiles):

- **Pfizer Posterior:**

  Posterior median: $\approx `r round(qbeta(0.5, 0.700102 + 8, 162 + 1), 4)`$

  95\% CI: $\Bigl[ `r round(qbeta(0.025, 0.700102 + 8, 162 + 1), 4)`,\; `r round(qbeta(0.975, 0.700102 + 8, 162 + 1), 4)` \Bigr]$

- **0.05 Prior Posterior:**

  Posterior median: $\approx `r round(qbeta(0.5, prior1Beta[1] + 8, 162 + prior1Beta[1]), 4)`$

  95\% CI: $\Bigl[ `r round(qbeta(0.025, prior1Beta[1] + 8, 162 + prior1Beta[1]), 4)`,\; `r round(qbeta(0.975, prior1Beta[1] + 8, 162 + prior1Beta[1]), 4)` \Bigr]$

- **0.01 Prior Posterior:**

  Posterior median: $\approx `r round(qbeta(0.5, prior2Beta[1] + 8, 162 + prior2Beta[1]), 4)`$

  95\% CI: $\Bigl[ `r round(qbeta(0.025, prior2Beta[1] + 8, 162 + prior2Beta[1]), 4)`,\; `r round(qbeta(0.975, prior2Beta[1] + 8, 162 + prior2Beta[1]), 4)` \Bigr]$
\section*{Hypothesis Tests}

We test the hypothesis for vaccine effectiveness:
\[
H_0: \pi \ge 0.5 \quad \text{vs} \quad H_1: \pi < 0.5.
\]

For each posterior distribution, the p-value is computed as the probability of observing 
\(\pi \ge 0.5\) under the posterior, i.e.,
\[
p\text{-value} = 1 - F(0.5),
\]
where \(F(\beta)\) is the cumulative distribution function of the corresponding Beta posterior.

\text{Pfizer Posterior:}
\[
p\text{-value} \approx `r round(1 - pbeta(0.5, 0.700102 + 8, 162 + 1), 4)`.
\]

\text{0.05 Prior Posterior:}
\[
p\text{-value} \approx `r round(1 - pbeta(0.5, prior1Beta[1] + 8, 162 + prior1Beta[1]), 4)`.
\]

\text{0.01 Prior Posterior:}
\[
p\text{-value} \approx `r round(1 - pbeta(0.5, prior2Beta[1] + 8, 162 + prior2Beta[1]), 4)`.
\]





# Discussion / Conclusion
Discuss / conclude here.

# Bibliography

Brown, B. (2024). *Lecture Title*. Lecture slides, Course Name, University Name.

Doe, J. (2020). Title of the Paper. *Journal Name*, 12(3), 45-67.

Last, F., & Last, F. (2025). *Book Title*. Publisher.

Smith, A., & Johnson, C. (2023). *Title of the Online Article*. Retrieved from https://www.example.com.

# Appendix

## Code
```{r ref.label = "important_R_code", eval=FALSE}
```

```{r ref.label = "psi_dist", eval=FALSE }
psi_dist <- function(x, shape1, shape2){
  pi <- (x-1)/(x-2)
  return(dbeta(x = pi, shape1 = shape1, shape2 = shape2))
}
```

## Plots
```{r, echo = FALSE}

ggplot() +
geom_function(fun = dbeta,
mapping = aes(color = "Pfizer"),
args = list(shape1 = 0.700102, shape2 = 1),
xlim = c(0,1)) +
geom_function(fun = dbeta,
mapping = aes(color = "Pfizer"),
args = list(shape1 = 0.700102 + 8, shape2 = 162 + 1),
xlim=c(0,1)) +
geom_function(fun = dbeta,
mapping = aes(color = "0.05"),
args = list(shape1 = 43.04, shape2 = 43.04),
xlim = c(0,1)) +
geom_function(fun = dbeta,
mapping = aes(color = "0.05"),
args = list(shape1 = 43.04 + 8, shape2 = 162 + 43.04),
xlim=c(0,1)) +
geom_function(fun = dbeta,
mapping = aes(color = "0.01"),
args = list(shape1 = 85.63, shape2 = 85.63),
xlim = c(0,1)) +
geom_function(fun = dbeta,
mapping = aes(color = "0.01"),
args = list(shape1 = 85.63 + 8, shape2 = 162 + 85.63),
xlim=c(0,1)) +
scale_color_manual(name = "dist", 
                   values = c("Pfizer" = "blue", "0.05" = "red", "0.01" = "black"))+
labs(title = "Priors and Posteriors",
x=expression(pi),
y = "PDF")


```

## Proofs
If applicable, include detailed mathematical derivations or additional theoretical explanations.

Proof of the Maximum Likelihood Estimator for $\widehat{\psi}_0^{mle}$:

We can begin with the Likelihood function. Recall that the design of our experiment is binomial; we observe 170 infections and try to make inference on $\pi$, the proportion of infections from those who have the vaccine. Assuming each observation is independent (which is necessary for a binomial model), we can write our likelihood function in terms of $\pi$:
$$L(\pi) = \binom{170}{x}(\pi)^x(1-\pi)^{170-x}$$
Where $x$ is the observed number of infections where the patient had the vaccine. We observe $x = 8$, so we can rewrite our likelihood function:
$$L(\pi; x = 8) = \binom{170}{8}(\pi)^8(1-\pi)^{162}$$
Notably, however, we are not hoping to draw inference on $\pi$, but rather on $\psi$, the efficacy of the vaccine. We can write this parameter in terms of $\pi$.
$$\psi = \frac{1-2\pi}{1-\pi}$$
In order to write our Likelihood function in terms of $\psi$, we can rewrite $\psi$ in terms of $\pi$:
$$\pi = \frac{\psi - 1}{\psi -2}$$
Now we rewrite our likelihood function.
$$L^*(\psi) = \binom{170}{8}\left(\frac{\psi-1}{\psi-2}\right)^8\left(\frac{-1}{\psi-2}\right)^{162}$$
Now that we have written our full likelihood function in terms of $\psi$, we can write the log likelihood function.
$$\ell^*(\psi) = \log\binom{170}{8}+8\log(\psi-1)-8\log(\psi-2)-162\log(\psi-2) +162\log(-1)$$
We have $162\log(-1) = \log((-1)^{162}) = \log(1) = 0$:
$$\ell^*(\psi) =  \log\binom{170}{8}+8\log(\psi-1)-170\log(\psi-2)$$
To maximize this equation, we take its derivative with respect to $\psi$:
$$\frac{d}{d\psi}\ell^*(\psi) = \frac{8}{\psi-1} - \frac{170}{\psi-2}$$
Setting this derivative finds the maximum:
$$0 = \frac{8}{\hat{\psi}_0^{mle} - 1} - \frac{170}{\hat{\psi}_0^{mle}-2}$$
$$\frac{170}{\hat{\psi}_0^{mle}-2} = \frac{8}{\hat{\psi}_0^{mle}-1}$$
$$170\hat{\psi}_0^{mle} - 170 = 8\hat{\psi}_0^{mle} - 16$$
$$\hat{\psi}_0^{mle} = \frac{154}{162}$$
